Income Has Serious Implications On Many Facets of American Life in 2021

### 2022-03-21 outputs/paper/paper.Rmd
9th commit

# Data

In this Data Section \@ref(data), we will provide a look into the data acquisition and processing methodology as well as a deep dive into the contents of the data. First of, (Table \@ref(tab:tableofdata)) give us a glimpse of the data.

```{r echo = F}
#table of variables that I am concerned with

reduced_data[complete.cases(reduced_data), ] %>%
   slice(1:7) %>%
  kable(
    caption = "GSS Cleaned Data",
    col.names = c("Row Number", "Age", "Gender", "Born", "Marital Status", "Education", 
                  "Divorce Status", "Degree", "Current Income", "Health Care", "Birth Year", 
                  "First Born", "Income at 16", "Grandparents", "Category for Income at 16", 
                  "Caegory for Current Income"),
    digits = 1,
    booktabs = TRUE,  
    linesep = "")
```

@@

# Visualizing the Data and The Implications
In this analysis, I am interested at looking at how real the American Dream is and if it is found to be possible, whether or not there is any difference in attainment amoung genders. As such, in this Section \@ref(visualizing-the-data-and-the-implications), we will create various visuals which will enable us to extract as much information as we can and we will speak about the implications that they make.
```{r, echo = F, message = F, include = F}
#creating levels for the x and y axis on the graphs

@@

income16_gender/income_gender
```
In figure \@ref(fig:figure-2), we see the overlapping of the two graphs, the first of which is the same as figure \@ref(fig:figure-1) and the bottom one is the same in essence but with the information for current family income instead of at age 16. The first and most striking part of this visual is that for the current income, the highest threshold is $25,000 or more. We see that the both of our graphs are heavily skewed to wards higher incomes. I suspect that this is mainly a production of the bin widths decided on for the analysis.

@@

## Implications 2
In the above visuals, we can see that in all the income groupings for both age 16 and current income, the largest subgrouping within each is people who have no grandparents born in the United states. The second largest group are usually those with 2 grandparents born in the United States. It is pretty logical that 2 grandparents follows after zero grandparents as it is easy to recognize that most people do not go abroad to marry and marry a friend or someone that they know locally, who is most likely to be born in the united states. We would speculate that a similar trend can be observed in other *immigrant* countries Like Canada or Sweeden. What is the most fascinating however is that the vast majority of people who answered the survey are first or second generation Americans, perhaps having grandparents or parents who immigrated to this country chasing the American Dream.

In a similar fashion to what was explained in \@ref(implications-1), both of our graphs are heavily skewed towards higher income. In the case of current income, it is almost certainly because of the bin width decided on by the people who built the survey which we personally find many problems with. In Section \@ref(limitations), we will take a deep dive into this.

@@

# Limitations
In this limitations section, I want to focus on the data which is provided to us and have a  look into why it might not be easy to take the implications mentioned above at face value

## Uneven Bin width for Income Groups.
The first major flaw I've found in the data is that the bins that both the income at age 16 and current income are recorded in are not all the same size. For example, in the case of the the first bin has a size of 16 years, the second bin from ages 16 to 24 has width 9, the next two have width 19 with the last ranging in theory from 65 to infinity. Although it is possible to understand why the maintainers of the data choose to have a demarcation point of the bins at 65 and 16, it is very hard to understand why they opted to divide the interior portion into varying bin sizes.

We know in data science that when plotting bar charts, the bin width can have a big implication on the presentation of the data and the things that we can infer about its distribution. If the bin width is too small or too large, we might reduce or add characteristics to the data that aren't actually representative of it.

## Why cut off income at $25,000?
use the us labor statistics data
## Third discussion point

>>>>

# Data
In this Data Section \@ref(data), we will provide a look into the data acquisition and processing methodology as well as a deep dive into the contents of the data. First of, (Table \@ref(tab:tableofdata)) give us a glimpse of the data.
```{r echo = F}
#table of variables that i am concerned with
data_table <- reduced_data[,-c(1,6,8)]

data_table[complete.cases(data_table), ] %>%
   slice(1:7) %>%
  kable(
    caption = "GSS Cleaned Data",
    col.names = c("Age", "Gender", "Born", 
                  "Education", 
                  "Birth Year", "Grandparents", 
                  "Income at 16","Current Income"),
    digits = 1,
    booktabs = TRUE,  
    linesep = "")

@@

# Visualizing the Data and The Implications
In this analysis, we am interested at looking at how real the American Dream is and if it is found to be possible, whether or not there is any difference in attainment amoung genders. As such, in this Section \@ref(visualizing-the-data-and-the-implications), we will create various visuals which will enable us to extract as much information as we can and we will speak about the implications that they make.
```{r, echo = F, message = F, include = F}
#creating levels for the x and y axis on the graphs

@@

income16_gender/income_gender
```
In figure \@ref(fig:figure-2), we see the overlapping of the two graphs, the first of which is the same as figure \@ref(fig:figure-1) and the bottom one is the same in essence but with the information for current family income instead of at age 16. The first and most striking part of this visual is that for the current income, the highest threshold is $25,000 or more. We see that the both of our graphs are heavily skewed to wards higher incomes.wesuspect that this is mainly a production of the bin widths decided on for the analysis.

@@

## Implications 2
In the above visuals, we can see that in all the income groupings for both age 16 and current income, the largest sub grouping within each is people who have no grandparents born in the United states. The second largest group are usually those with 2 grandparents born in the United States. It is pretty logical that 2 grandparents follows after zero grandparents as it is easy to recognize that most people do not go abroad to marry and marry a friend or someone that they know locally, who is most likely to be born in the United States. We would speculate that a similar trend can be observed in other *immigrant* countries like Canada or Sweden. What is the most fascinating however, is that the vast majority of people who answered the survey are first or second generation Americans, perhaps having grandparents or parents who immigrated to this country chasing the American Dream.

In a similar fashion to what was explained in Section \@ref(implications-1), both of our graphs are heavily skewed towards higher income. In the case of current income, it is almost certainly because of the bin width decided on by the people who built the survey which we personally find many problems with. In Section \@ref(limitations), we will take a deep dive into this.

@@

# Limitations
In this limitations section,wewant to focus on the data which is provided to us and have a  look into why it might not be easy to take the implications mentioned above at face value.

## Uneven Bin width for Income Groups.
The first major flaw I've found in the data is that the bins that both the income at age 16 and current income are recorded in are not all the same size. For example, in the case of the current income variable, the the first bin has a size of \$1,000, the second bin has a width of \$1,999, the third returns to about \$1,000 and as we go higher up the echelon, we have seemingly random alterations in bin widths from \$1,000 to \$2,000 to \$5,000. This can cause serious problems in statistics and especially for purposes of analysis.

We know in data science that when plotting bar charts, the bin width can have a big implications on the presentation of the data and the things that we can infer about its distribution. If the bin width is too small, too large or uneven, we might reduce or add characteristics to the data that aren't actually representative of it. This will cause us to over-fit or under-fit a model when the time comes to create them. The other problem with the bin widths is that they are different from the income at age 16 to the current income. This is the subject of the following Section 5.2.

## Why cut off income at $25,000?
According to a news release published by the United States Bureau of Labor Statistics (see the document called wkyeng.pdf in the literature folder in inputs), the median weekly income of US citizens in the 4th quarter of 2021 (furthest from the start of the pandemic) is $1,010. This information is from that of 116.3 million full-time wage and salary workers. If we do some basic mathematics to approximate the annual earning, we get a figure of over \$52,000 dollars per year. As such, we are unable to understand why the bins for current income have been designed as such. This is a major flaw in the technique used by the surveyers to collect the data. In section \@ref(next-steps)
